Systematically Optimizing Abstract Abstract Machines

### 2013-03-29 paper/paper.tex
Added changes we just discussed.

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\section{Reduction semantics to baseline analyzer}
\label{sec:baseline}

@@

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\\
\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
We associate the value in hand with the expression that produced
it. Also, we fan out to all functions that are associated with the
original function position, because applied functions are in
\emph{strict position} (more about that later).

@@

\begin{align*}
\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont} & \dmachstep_\msto
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(\mastate',\msdiff) \mid \mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
 S' &= \setof{(\mastate,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

@@

\begin{lemma}
$\mastate \dmachstep_\msto \ttuple{\mastate'}{\msdiff}$ iff
$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

@@

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We provide a formal
analysis of abstract compilation \emph{without} store widening with a proof
of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
semantics and lazy non-determinism without widening to show precision
preservation. We revisit store widening at the end of the section.

@@

\begin{theorem}[Precision preservation]
$B$ is a WEB on $\lmachstep \cup \machstep$
\end{theorem}

@@

\begin{enumerate}
\item{$S$: the \emph{seen} set, though made a map for faster membership tests and updates.}
\item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
\item{$\msto$: the store, which represents all stores that occur in the reduction semantics.}
\item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
\item{$\changep$: whether there has been a store change stepping all states in $F$.}
\end{enumerate}

>>>>

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\maddr \mbox{ where } \maddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses. The $\sqcup$ on
stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
relation is non-deterministic in its choice of addresses, however it
must always choose a fresh address when allocating a continuation or
variable binding.  If we consider machine states equivalent up to
consistent renaming and fix an allocation scheme, this relation
defines a deterministic machine (the relation is really a function).

@@

\section{Machine semantics to baseline analyzer}
\label{sec:baseline}

@@

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision.

\begin{align*}
\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is demanded in order to continue with
evaluation. More generally, this is because applied functions are in
\emph{strict position} (more about that later).

@@

\begin{align*}
\dmachstep &\subseteq (\widehat{\State}\times\Store) \times (\widehat{\State}\times\StoreDelta) \\
\ttuple{\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\mkont}}{\msto} & \dmachstep
\ttuple{\ev[^{\mcntr'}]{\mexp,\menv',\mkont}}{\cons{\ttuple{\maddr}{\setof{\mval}}}{\epsilon}} \\
\text{ where }\maddr &= \alloc(\mstate) \\
              \menv' &= \menv[\mvar\mapsto\maddr]
\end{align*}

@@

\begin{align*}
(S, F, \msto,\Sigma,t) &\mathrel{\damachstep} (S \cup S', F', \msto',\Sigma',t') \\
\mbox{ where }
 I &= \setof{(\mastate',\msdiff) \mid \ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff} } \\
 \ttuple{\msto'}{\changep} &= \replay(\appendall(\setof{\msdiff \mid (\_,\msdiff) \in I}),\msto) \\
 \ttuple{t'}{\Sigma'} &=
     \left\{
       \begin{array}{ll}
         \ttuple{t+1}{\msto\Sigma} & \text{ if } \changep \\
         \ttuple{t}{\Sigma} & \text{ otherwise}
       \end{array}\right. \\
 F' &= \setof{\mastate \mid (\mastate,\_) \in I, (\mastate,t') \notin S} \\
 S' &= \setof{(\mastate,t') \mid c \in F'} \\
\appendall(\varnothing) &= \epsilon \\
\appendall(\setof{\msdiff}\cup\Xi) &= \append(\msdiff,\appendall(\Xi))
\end{align*}

@@

\begin{lemma}
$\ttuple{\mastate}{\msto} \dmachstep \ttuple{\mastate'}{\msdiff}$ iff
$\wn(\mastate,\msto) \machstep \wn(\mastate',\replay(\msdiff,\msto))$
\end{lemma}
By cases on $\dmachstep$ and $\machstep$.

@@

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We provide a formal
analysis of abstract compilation \emph{without} store widening with a proof
of a stuttering bisimulation~\cite{ianjohnson:BCG88} between this
semantics and lazy non-determinism without widening to show precision
preservation. %We revisit store widening at the end of the section.

@@

\begin{theorem}[Precision preservation]
$B$ is a WEB on $\lmachstep \uplus \machstep$
\end{theorem}

@@

\begin{enumerate}
\item{$S$: the \emph{seen} set, though made a map for faster membership tests and updates.}
\item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
\item{$\msto$: the store, which represents all stores that occur in the machine semantics.}
\item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
\item{$\changep$: whether there has been a store change stepping all states in $F$.}
\end{enumerate}
